\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-02 chapters/gaze_independence.tex
Comments Marc

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
	\caption{Incidence of eye motor impairment in selected \ac{bci} target
    patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
    \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
    $+$: frequent, $-$: infrequent.}
    \label{tab:incidence}
\end{table}
\todo{include FA}
\todo{cite Abnormal Oculomotor Functions in Amyotrophic Lateral Sclerosis 10.3988/jcn.2018.14.4.464}
\todo{cite Neuro-Ophthalmologic Manifestations of Amyotrophic Lateral Sclerosis}

@@

\section{Gaze-independent \ac{bci} paradigms}
\label{sec:gaze-independence}
\todohsm{rename to state of the art in gaze-independent bci. Address more systematically.}

Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and
somatosensory\todo{cite some}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
Furthermore, they can suffer from increased mental effort in operation and
user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
speller~\parencite{Acqualagna2011}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific \ac{erp} components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
Nevertheless, these methods can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, these systems typically require users to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.
\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}

Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.
More recent work has made advances in decoding lateralized covert
VSA harnessing the N2pc \ac{erp} component~\cite{Thiery2016,Reichert2020b,Wang2022},
decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
stimulation paradigms~\cite{Egan2017}.
Since these methods often require a slower stimulation pace than the classical
\ac{erp} \ac{bci} paradigm, gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

The general goal of this PhD is to tackle the gaze-dependence in visual \ac{bci}s.
\todo{refer to ITRs from previous chapter}
Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like Rapid Serial Visual
Presentation\todo{cite},
where targets or small sets of targets are shown consecutively.
Traditionally, spatially organized interfaces have a higher information transfer
rate, but are also more gaze-dependent than temporally organized interfaces.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

@@

The interface can be operated in different VSA conditions as illustrated in
Figure~\ref{fig:gaze-vsa}.
\fourpanefig{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
  }{%
    \emph{Overt \ac{vsa}}.
    Gaze and \ac{vsa} coincide on a target.
  }{%
    fig:gaze/vsa/overt
  }
}{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
  }{%
    \emph{Covert \ac{vsa}}.
    Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
  }{%
    fig:gaze/vsa/covert
  }
}{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
  }{%
    \emph{Split \ac{vsa}}.
    \Ac{vsa} is directed towards a target, while the gaze rests on another.
  }{%
    fig:gaze/vsa/split
  }
}{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
  }{%
    \emph{Free \ac{vsa}}.
    The user is free to direct their gaze as they deem most comfortable
    All previous \ac{vsa} conditions are possible.
  }{%
    fig:gaze/vsa/free
  }
}{%
  \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
}{%
  fig:gaze-vsa
}

In the overt case, users gaze at the cued
target they are also mentally attending; in the covert case, users
gaze at the center of the screen while mentally attending to the cued target.
Finally, we introduce a VSA condition that is understudied in the context of
gaze-independent \ac{bci} development: split VSA.
In split VSA, the user mentally focuses on one cued target while gazing at
another (the distractor).
\todomvh{+ "free" condition in Fig. 21.}

>>>>

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
  \caption[Incidence of eye motor impairment in selected \ac{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}
\todo{include FA}
\todo{cite Abnormal Oculomotor Functions in Amyotrophic Lateral Sclerosis 10.3988/jcn.2018.14.4.464}
\todo{cite Neuro-Ophthalmologic Manifestations of Amyotrophic Lateral Sclerosis}

@@

\section{Gaze-independent \ac{bci} paradigms}
\label{sec:gaze-independence}
\todohsm{rename to state of the art in gaze-independent bci. Address more systematically.}
\todo{in a paragraph describe each approach and its up and downsides}
\todo{move form copied introduction}

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and
somatosensory\todo{cite some}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
Furthermore, they can suffer from increased mental effort in operation and
user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

\subsection{Gaze-independent visual stimulation}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific \ac{erp} components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
Nevertheless, these methods can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, these systems typically require users to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.
\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}

\subsection{Gaze-independent decoding}
Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.
More recent work has made advances in decoding lateralized covert
VSA harnessing the N2pc \ac{erp} component~\cite{Thiery2016,Reichert2020b,Wang2022},
decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
stimulation paradigms~\cite{Egan2017}.
Since these methods often require a slower stimulation pace than the classical
\ac{erp} \ac{bci} paradigm, gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

The general goal of this PhD is to tackle the gaze-dependence in visual \ac{bci}s.
\todo{refer to ITRs from previous chapter}
Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}\todo{cite},
where targets or small sets of targets are shown consecutively.
Traditionally, spatially organized interfaces have a higher information transfer
rate, but are also more gaze-dependent than temporally organized interfaces.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

@@

The interface can be operated in different VSA conditions as illustrated in
Figure~\ref{fig:gaze-vsa}.

\begin{figure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
    \caption[Overt \ac{vsa}]{%
      \emph{overt} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
    \caption[Covert \ac{vsa}]{%
      \emph{Covert} \ac{vsa}.
      Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
    }
    \label{fig:gaze/vsa/covert}
  \end{subfigure}

  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
   \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
      The user is free to direct their gaze as they deem most comfortable
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze-vsa}
\end{figure}
In the overt case, users gaze at the cued
target they are also mentally attending; in the covert case, users
gaze at the center of the screen while mentally attending to the cued target.
Finally, we introduce a VSA condition that is understudied in the context of
gaze-independent \ac{bci} development: split VSA.
In split VSA, the user mentally focuses on one cued target while gazing at
another (the distractor).
\todomvh{+ "free" condition in Fig. 21.}

### 2024-10-02 chapters/patients.tex
Comments Marc

\section{Introduction}
\todo{rename patient to bci user and participant}

BCIs for communication assistive technology~\cite{Millan2010}
have a target population consisting of patients people with \ac{sspi}.

@@

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-independent solutions.
These studies build on the assumption that patients with eye motor impairment
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when a central
fixation is maintained, could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors due to their eye
motility or vision.

@@

\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed a trial with a
interface with two participants with late-stage \ac{als} and visual impairment.
Their system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be verified with other
patient groups and visual oddball \acp{bci}.

\textcite{Severens2014} evaluated the visual Hex-o-Spell~\cite{Treder2010} on 5
participants with \ac{als} and showed that this visual oddball interface optimized
for gaze-independence can outperform a tactile gaze-independent \ac{bci}.
While this speaks to the power of visual paradigms even in groups that are
expected to have eye motor impairment, they did not verify the gaze direction
of participants during the experiment.
It was suggested that patients were performing overtly.
Participants with \ac{als} also had a substantially lower accuracy patients lower than healthy
(58\% vs 88\%).

Our previous study, presented in chapter \todo{ref chapter} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that gaze impaired patients might
not full rely on central gaze fixation and evaluates settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved
improved in healthy subjects using a suited decoding strategy.
that accounts for latency jitter in covert \ac{vsa} responses improves
Yet, there is a strong need for verification of these result in an applied
setting with people with \ac{sspi}.

@@

\fourpanefig{%
  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}}{A
  patient with the stimulation and recording setup.}{fig:patients/photos/side}
}{%
  \figpane{\includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}}{subcaption}{fig:patients/photos/front}
}{%
  the interface
}{%
  4
}{%
  caption
}{fig:patients/photos}

@@

In overt \ac{vsa}, evaluated decoders performed about equal on average
(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, tLDA 75.99\%).
In covert \ac{vsa} with cued central gaze fixation, performance deteriorates
but \ac{wcble} can significantly improve performance over its base classifier
\ac{tlda} in this condition
(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, tLDA 59.05\%).
Decoding performance for this task was at chance level for participants PB4 and
PC3.

However, \Ac{wcble} did not improve \ac{tlda} in free \ac{vsa} free \ac{vsa}, but
XDAWNCov+TS+LDA performance was slightly lower here (although not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, tLDA 74.27\%).
More interestingly, we notice that performances of gaze impaired individuals
with \ac{sspi} in free \ac{vsa} are close to those in overt \ac{vsa}.
A substantial decrease in performance from overt to free is present for
subjects PC3 and PC4.
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
setting, there was no decrease in performance.

@@

Due to the heterogenous nature of the participant's conditions, it is hard to support
claims about significance of presented results.
This study should therefore be seen as a collection of case studies,
highlighting different obstacles encountered in developing gaze-independent
visual oddball \acp{bci} for individuals with \ac{sspi} and gaze impairment.
Nevertheless, we would like to highlight some general aspects that might be
of interest for the further development of this class of \acp{bci}.

True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
Hence our analysis centers around the free \ac{vsa} condition.
Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
confirm our assumption that voluntary covert \ac{vsa} can
occur in individuals with \ac{sspi} and gaze impairment.
We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
chapter\todo{ref} that state that decoding of covert \ac{vsa} with central gaze
fixation can be improved by accounting for latency jitter, and showed that this
holds for individuals with \ac{sspi} and gaze impairment.

Contradicting our assumptions, however, we have shown that this does not
necessarily improve covert \ac{vsa} when gaze fixation is not cued.
One possible explanation for this is that actively performing central gaze fixation
increases task load.
This in turn reduced overall performance, even though the participant might
otherwise also have performed covert \ac{vsa}, but would not be occupied by
maintaining strict central gaze fixation.
These extended task demands are not present in the free \ac{vsa} condition, so
there is less performance to be gained here.
Furthermore cued central gaze fixation combined with counting flashing stimuli
in the visual periphery, is an explicit example of a dual task.
Dual tasks have been shown to increase P3 latency
jitter~\cite{Polich2007,Arico2014, VanDenKerchove2024}
which is what \ac{wcble} accounts for.
Hence,increased P3 jitter might be more related to maintaining central gaze fixation
than to the actual covert \ac{vsa} aspect.

@@

Individuals with all but the most severe gaze impairments will probably retain
some degree of gaze direction in visual \ac{bci} operation, which can
drastically boost performance.
Subject PB2 exemplifies this: his free \ac{vsa} performance is on par with
This is also supported by our results on cross contidion calibration

@@

Another limiting factor are the difficulties in correctly interpreting eye tracker
results in studies with eye motor impaired patients.
If eye tracking at all is possible, it is not guaranteed that the user is able
to perform the calibration procedure succesfully.
Further experiments should be carried out with a stationary eye tracker with
more advanced capabilities, although systems using a head fixator or headrest
should be avoided.
This is not practical when working with
wheelchair-bound individuals who might have undergone a tracheostomy.
hard because of inaccuracy eye tracker for eye
motor impaired patients. Redo with more advanced stationary eye tracker and
head rest.

@@

To summarize, we present our three main findings.
\begin{enumerate*}
  \item We have confirmed that some individuals with \ac{sspi} and gaze
    impairment find it more comfortable to operate a visual gaze-independent
    \ac{bci} relying on covert \acf{vsa}.
  \item We have shown that covert \acf{vsa} with central gaze fixation can be
    improved by a suitable decoding strategy, namely the \acf{wcble} decoder,
    but this does not hold true if central gaze fixation is not imposed.
  \item We have shown that covert \ac{vsa} \todo{with or
    without imposed gaze fixation} decoding can alternatively be improved by
    using a calibration session with attempted overt \ac{vsa}.
\end{enumerate*}

>>>>

\section{Introduction}
\todo{rename patient to bci user and participant}
\todomvh{separate terminology for sspgi}
\todo{consistently: gaze impairment, eye motor impairment or oculomotor impairment}

\todo{A systematic review of research on augmentative and alternative communication brain-computer interface systems for individuals with disabilities}

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \ac{sspi}.

visual bci

@@

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-independent solutions.
These studies build on the assumption that patients with eye motor impairment
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when central
fixation is maintained, could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors resulting from their eye
motility.

@@

\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed an trial with an
two participants with late-stage \ac{als} and visual impairment.
Their \ac{ssvep} interface was not optimized for gaze-independence, but the
system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be verified with other
patient groups and visual oddball \acp{bci}.

\textcite{Orhan2012} and \textcite{Oken2014} tested the \ac{rsvp} speller with
individuals with \ac{lis}

\textcite{Severens2014} evaluated the visual Hex-o-Spell~\cite{Treder2010} on 5
participants with \ac{als} and showed that this visual oddball interface optimized
for gaze-independence can outperform a tactile \ac{bci}.
While this speaks to the power of visual paradigms even in groups that are
expected to have eye motor impairment, they did not verify the gaze direction
of participants during the experiment.
It was suggested that patients were performing overtly.
Participants with \ac{als} also had a substantially lower accuracy than healthy
controls (58\% vs 88\%).

Our previous study, presented in chapter \todo{ref chapter} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that gaze impaired patients might
not fully rely on central gaze fixation and evaluates settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved
improved in healthy subjects using a suited decoding strategy.
that accounts for latency jitter in covert \ac{vsa} responses improves
Yet, there is a strong need for verification of these result in an applied
setting with people with \ac{sspi}.

@@

\begin{figure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}
    \caption{A patient with the stimulation and recording setup.}
    \label{fig:patients/photos/side}
  \end{subfigure}\hfill%
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}
    \caption{caption}
    \label{fig:patients/photos/front}
  \end{subfigure}
  \caption{The recording setup and stimulation interface.}
  \label{fig:patients/photos}
\end{figure}
\todo{caption}
\todo{include stimulation interface}

@@

In overt \ac{vsa}\todomvh{specify setting when talking about the setting}, evaluated decoders performed about equal on average
(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
In covert \ac{vsa} with cued central gaze fixation, performance deteriorates
but \ac{wcble} can significantly improve performance over its base classifier
\ac{tlda} in this condition
(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, \ac{tlda} 59.05\%).
Decoding performance for this task was at chance level for participants PB4 and
PC3.

However, \Ac{wcble} did not improve \ac{tlda} in free \ac{vsa} free \ac{vsa}, but
XDAWNCov+TS+LDA performance was slightly lower here (although not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
More interestingly, we notice that performances of the decoders in free
\ac{vsa} are close to those in overt \ac{vsa}.
A substantial decrease in performance from overt to free is present for
subjects PC3 and PC4.
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
setting, there was no decrease in performance.

@@

Due to the heterogenous nature of the participant's conditions, it is hard to support
claims about significance of presented results.\todomvh{
weird formulation
hard to draw general conclusions}
This study should therefore be seen as a collection of case studies,
highlighting different obstacles encountered in developing gaze-independent
visual oddball \acp{bci} for individuals with \ac{sspi} and gaze impairment.
Nevertheless, we would like to highlight some general aspects that might be
of interest for the further development of this class of \acp{bci}.

True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.\todomvh{any reference/evidence in support of this statement?}
Hence our analysis centers around the free \ac{vsa} condition.
Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
confirm our assumption that voluntary covert \ac{vsa} can
occur in individuals with \ac{sspi} and gaze impairment.
We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
chapter\todo{ref} that state that decoding of covert \ac{vsa} with central gaze
fixation can be improved by accounting for latency jitter, and showed that this
holds for individuals with \ac{sspi} and gaze impairment.

Contradicting our assumptions, however, we have shown that this does not
necessarily improve covert \ac{vsa} when gaze fixation is not cued.
One possible explanation for this is that actively performing central gaze fixation
increases task load.
This can in turn reduce overall performance, even though the participant might
otherwise also have performed covert \ac{vsa}, but would not be occupied by
maintaining strict central gaze fixation.
These extra task demand is not present in the free \ac{vsa} condition, so
there is less performance to be gained here.
Furthermore cued central gaze fixation combined with counting flashing stimuli
in the visual periphery, is an explicit example of a dual task.
Dual tasks have been shown to increase P3 latency
jitter~\cite{Polich2007,Arico2014, VanDenKerchove2024}
which is what \ac{wcble} accounts for.
Hence,increased P3 jitter might be more related to maintaining central gaze fixation
than to the actual covert \ac{vsa} aspect.

@@

Individuals with all but the most severe gaze impairments will probably retain
some degree of gaze direction in visual \ac{bci} operation, which can
drastically boost performance.
Subject PB2 exemplifies this: his free \ac{vsa} performance is on par with his
overt \ac{vsa} performance, although eye tracking showed that he relied mostly
on over \ac{vsa} when cued to do so, and mostly on covert \ac{vsa} when
gaze was uncued.
This is also supported by our results on cross condition calibration\todo{how}

@@

Another limiting factor are the difficulties in correctly interpreting eye
tracker results in studies with eye motor impaired patients.
If eye tracking at all is possible, it is not guaranteed that the user is able
to perform the calibration procedure successfully.
Further experiments should be carried out with a stationary eye tracker with
more advanced capabilities, although systems using a head fixator or headrest
should be avoided.
This is not practical when working with
wheelchair-bound individuals who might have undergone a tracheostomy and might
suffer from spasticity.

@@

To summarize, we present our three main findings.
\begin{enumerate*}
  \item We have confirmed that some individuals with \ac{sspi} and gaze
    impairment find it more comfortable to operate a visual gaze-independent
    \ac{bci} relying on covert \acf{vsa}.
  \item We have shown that covert \acf{vsa} with central gaze fixation can be
    improved by a suitable decoding strategy, namely the \acf{wcble} decoder,
    but this does not hold true if central gaze fixation is not imposed.
  \item We have shown that covert \ac{vsa}
    decoding can alternatively be improved by
    using a calibration session with attempted overt \ac{vsa}.
\end{enumerate*} \todo{with or without imposed gaze fixation}
